---
title: "Nonconsulting writing: Logistic Regression"
output:
  html_document:
    css: styles.css
    toc: true
    toc_depth: 2
    number_sections: true
    df_print: paged
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5
)
```

```{r}
library(xtable)
library(tidyverse)
library(pROC)
library(ggplot2)
```


# Bernoulli and Binomial random variables 

R code simulating Bernoulli (0/1) random outcomes:
```{r}
p1 <- 0.5
yi <- rbinom(5, 1, p1); yi
```

R code simulating Binomial random outcomes:
```{r}
p2 <- 0.7
yi <- rbinom(10, 5, p2); yi
```


Plot for binary response variable compared to continuous response variable
```{r}
# ---- Simulate two types of responses ----
n <- 100

# Continuous response (e.g. blood pressure, weight)
y_cont <- rnorm(n, mean = 50, sd = 10)

# Binary response (e.g. disease: 0 = no, 1 = yes)
p <- 0.3                  # probability of "1"
y_bin <- rbinom(n, 1, p)  # 0/1 outcomes

# ---- Plot: continuous vs binary response ----
par(mfrow = c(1, 2))

# Left: continuous response
plot(1:n, y_cont,
     pch = 19,
     xlab = "Observation index",
     ylab = "Continuous response",
     main = "Continuous response")

# Right: binary response
plot(1:n, jitter(y_bin, amount = 0.05),
     pch = 19,
     xlab = "Observation index",
     ylab = "Binary response (0 or 1)",
     main = "Binary response",
     yaxt = "n")
axis(2, at = c(0, 1), labels = c("0", "1"))
abline(h = 0:1, lty = 2, col = "grey")
par(mfrow = c(1, 1))
```


Running logistic regression with simulated data and interpretation on the coefficients.

The logistic regression model examining the relationship between $x$ and the binary outcome $y$ shows that $x$ is a statistically significant predictor with $z = 4.61$ and $p < 0.001$. While the estimated coefficient for $x$ is $0.134$, which corresponds to an odds ratio of $1.14$, this means that for every one-unit increase in $x$, the odds of the outcome being a `success` (1) increase approximately 14%, holding all else constant. The intercept estimate is $-6.54$ (odds ratio $\approx 0.0014$), representing the log-odds of the outcome being 1 when $x = 0$. 

Model fit statistics indicate improvement over the null model: the null deviance reduced from $165.15$ to $133.34$, meaning the model explains more variation in the outcome than a model with no predictors. The AIC value of $137.34$ can be used to compare this model with alternative models (lower AIC indicates better fit).
Overall, the results suggest strong evidence that higher values of $x$ are associated with a greater probability of $y = 1$. 
```{r}
# Simulated binary response
n <- 120
x <- rnorm(n, 50, 10)
p <- plogis((x - 50)/8)
y <- rbinom(n, 1, p)

# Logistic regression
logit_model <- glm(y ~ x, family = binomial)

summary(logit_model)
exp(coef(logit_model))   # Odds ratios
```


Predict probabilities of each observation has outcome = 1, can be derived from using the fitted logistic regression. 
```{r}
pred_prob <- predict(logit_model, type = "response")
head(pred_prob)
```

We turn probabilities into actual predicted classes (0 or 1). We choose a cutoff (threshold) value, typically 0.5--if probability $> 0.5$, then predict as 1, if probability $\leq 0.5$ then predict the outcome as 0--for constructing the classification table. The following printed table is called the confusion matrix, where it shows how many cases the model predicted correctly or incorrectly.
```{r}
threshold <- 0.5
pred_class <- ifelse(pred_prob > threshold, 1, 0)

table(Actual = y, Predicted = pred_class)
```

The ROC curve helps us evaluate how well the model classifies 0 vs 1 across all possible thresholds, not just 0.5. When the AUC = 0.5, it means that the model is no better than guessing, and AUC closer to 1 indicates a better classification. 
```{r}
roc_obj <- roc(y, pred_prob)
plot(roc_obj, col="darkgreen", lwd=3, main="ROC Curve")
auc(roc_obj)
```



# Data analysis with logistic regression

We us the `mtcars` dataset, which contains information of 32 car models. For logistic regression, we create a data frame with the following key variables to be included in the model. \ 

`am` is a variable indicating transmission type, which is our binary outcome. \
`wt` is car weight, a predictor measured in 1000 pounds. \
`hp` is horsepower, which is another continuous predictor. \
```{r}
data(mtcars)
df <- within(mtcars, {
  am <- as.integer(am)        # ensure binary 0/1
  wt <- wt                    # weight (1000 lbs)
  hp <- hp                    # horsepower
})

dim(mtcars)
```

The logistic regression model examines whether car weight `wt` and horsepower `hp` predict the probability of a car having a manual transmission `am` = 1, instead of automatic. 

Both predictors are statistically significant:  \

`wt` has a negative effect with estimate = $-8.08$, $p = 0.008$, indicating heavier cars are less likely to have manual transmission. \ 

`hp` has a positive effect with estimate = $0.036$, $p = 0.041$, indicating cars with higher horsepower are more likely to have manual transmission. \

```{r}
# 2) Fit logistic regression ---------------------------------
# logit(P(am=1)) = beta0 + beta1*wt + beta2*hp
fit <- glm(am ~ wt + hp, data = df, family = binomial(link = "logit"))

cat("\n=== Model summary (logit scale) ===\n")
print(summary(fit))
```

Additional interpretation on the coefficients. 
```{r}
# 3) Odds ratios and 95% CI ----------------------------------
# OR = exp(beta). For a one-unit increase in the predictor, odds multiply by OR.
coefs <- coef(fit)
OR <- exp(coefs); OR
```


We can also derive the confidence interval of the OR 
```{r}
# Profile-likelihood CI for coefficients, then exponentiate to get OR CI
ci_beta <- suppressMessages(confint(fit))
OR_CI <- exp(ci_beta)

cat("\n=== Odds Ratios (with 95% CI) ===\n")
print(data.frame(
  term = names(OR),
  OR = OR,
  OR_low = OR_CI[, 1],
  OR_high = OR_CI[, 2],
  row.names = NULL
), digits = 3)
```

Predicted probabilities for `am` = 1 for each observation 
```{r}
# 4) Predicted probabilities for each observation -------------
df$phat <- predict(fit, type = "response")  # P(am=1 | wt, hp)

cat("\n=== First 6 predicted probabilities (P(manual)) ===\n")
print(head(df[, c("wt", "hp", "am", "phat")], 6), digits = 3)
```



# Comparison of different link functions

In logistic regression and general linear models (GLMs) for binary outcomes, we do not model the probability directly. Instead, we model a transformed version of the probability using a link funciton. The code below plots the inverse link functions, showing how each link converts the linear predictor into a probability between 0 and 1. The x-axis represents the linear predictor 
\[ \eta = \beta_0 + \beta_1 x\]
and the y-axis shows the corresponding predicted probability $p$. 

```{r}
eta <- seq(-5, 5, length.out = 400)
curve_logit <- plogis(eta)
curve_probit <- pnorm(eta)
curve_cloglog <- 1 - exp(-exp(eta))
curve_cauchit <- pcauchy(eta)

plot(eta, curve_logit, type="l", lwd=2, xlab=expression(eta),
ylab=expression(p==g^{-1}(eta)), main="Inverse links: p(eta)")
lines(eta, curve_probit, lwd=2, lty=2)
lines(eta, curve_cloglog, lwd=2, lty=3)
lines(eta, curve_cauchit, lwd=2, lty=4)
legend("topleft", lwd=2, lty=1:4, bty="n",
legend=c("logit","probit","cloglog","cauchit"))

```


```{r}
lr.logit <- glm(am ~ wt + hp, family = binomial(link = "logit"), data = df)   
lr.probit <- glm(am ~ wt + hp, family = binomial(link = "probit"), data = df)  
lr.cloglog <- glm(am ~ wt + hp, family = binomial(link = "cloglog"), data = df) 
lr.cauchit <- glm(am ~ wt + hp, family = binomial(link = "cauchit"), data = df) 
```


